# Visualizing the Effects of Data Transformations on Errors

Robert M Flight and Hunter NB Moseley*

Department of Molecular and Cellular Biochemistry, Markey Cancer Center, 
Resource Center for Stable Isotope-Resolved Metabolomics, University of Kentucky, 
Lexington, KY, 40536, USA

*hunter.moseley@uky.edu

**Background**

In many omics analyses, a primary step in the data analysis is the application 
of a transformation to the data.  Transformations are generally employed to 
convert proportional error (variance) to additive error, which most statistical 
methods appropriately handle. However, omics data frequently contain error 
sources that result in both additive and proportional errors. To our knowledge, 
there has not been a systematic study on detecting the presence of proportional 
error in omics data, or the effect of transformations on the error structure. 

**Results**

In this work, we demonstrate a set of three simple graphs which facilitate the 
detection of proportional and mixed error in omics data when multiple replicates
are available. The three graphs illustrate proportional and mixed error in a 
visually compelling manner that is both straight-forward to recognize and to 
communicate. The graphs plot the 1) absolute difference in the range, 2)
standard deviation and 3) relative standard deviation against the mean signal
across replicates (Figure 1). In addition to showing the presence of different types of
error, these graphs readily demonstrate the effect of various transformations on
the error structure as well.

**Conclusions**

We have developed a straight-forward set of graphs that effectively summarize 
major types of error present in high replicate datasets.Using these graphical 
summaries we find that the log-transform is the most effective method of the
common methods employed for removing proportional error.

**Keywords**

error analysis, error visualization, omics error

```{r insert_figure, echo=FALSE, fig.width = 7, fig.height = 2.75}
library(fakeDataWithError)
library(visualizationQualityControl)
library(cowplot)
library(dplyr)
library(errorTransformation)

set.seed(1234)
n_point <- 1000
n_rep <- 100
offset <- 4
simulated_data <- c(rlnorm(n_point / 2, meanlog = 1, sdlog = 1),
                    runif(n_point / 2, 5, 100))
simulated_data <- simulated_data + offset

prop_sd <- 0.1
add_sd <- 0.5
mix_error <- add_prop_uniform(n_rep, simulated_data, add_sd, prop_sd)

me_nt_diff <- plot_diff(mix_error, use_title = NULL)
me_nt_sd <- plot_sd(mix_error, use_title = NULL) + geom_hline(yintercept = add_sd, color = "red")
me_nt_rsd <- plot_sd(mix_error, use_title = NULL, sd_type = "rsd") + 
  geom_hline(yintercept = prop_sd, color = "red")

plot_grid(me_nt_diff, me_nt_sd, me_nt_rsd, nrow = 1, labels = LETTERS[1:3])
```

**Figure 1**: Three plots summarizing the error in simulated data of 10000
features across 100 replicates, with an additive error of 0.5 and proportional
error of 0.1. The *log(mean)* is the *log* of the mean value across the 100
replicates for a feature. Red lines indicate the level of additive (A) and 
proportional (B) error added. **A**) The absolute difference between the highest and
lowest valued replicates vs the mean; **B**) the standard deviation (sd) vs the
mean; **C**) the relative standard deviation (rsd) vs the mean.

```{r runit, include=FALSE, eval=FALSE}
rmarkdown::render("vignettes/kbrin_poster_abstract.Rmd", 
                  output_format = c("word_document", "pdf_document"),
                  output_dir = "outputs")
```

